Zohmg---a large scale data store for aggregated time-series-based data. [DRAFT]

### 2009-05-12 doc/report/msc-report.tex
orphan'd text at the end.
doc/report/msc-report.tex: * general work on the Method section. * added to Glossary appendix.
doc/report/references.bib:
* removed Java.
* added OLAP Data Scalability

\newcommand{\chapterquote}[2] {
\begin{quote}
\textit{"{#1}"}

@@

At last.fm there is a central repository where metrics and statistics for
various services are stored. There is a web site for internal use that
displays graphs of the services and their statistics. Examples of statistics
are web site usage broken down by country, scrobbling broken down by client
and daily subscription counts.

@@

@@

@@

@@

\section{Interface}

@@

The slice operation fixes the values of one or several dimensions in the cube.
The data omitted from this operation would be any data associated with the
values of the dimension that were not fixed. The dice operation is a slice on
more than two dimensions. [source:
http://www.cs.ualberta.ca/~zaiane/courses/cmput690/glossary.html]

@@

Time is the most common of the dimensions in the data cube constructed for
Zohmg. The time stamp is a common property for all data handled by Zohmg;
every measurement is expected to have time as one of its dimensions. TODO:
really, this belongs in the implementation section. I find nothing interesting
to say about time series in this section.

@@

\section{System}

Regarding the infrastructure, both BigTable and MapReduce has high
availability, scalability and reliability. Since BigTable runs on top of
GFS\footnote{Google File System}, the data stored in BigTable is
replicated, usually three times, onto other nodes in the cluster.
\cite{gfs} Likewise, if a node in a MapReduce job goes down or fails to
complete the job it is reassigned to another node. \cite{mapreduce}

Hadoop is a free software implementation of Google's MapReduce-framework
described in \cite{mapreduce}. Hadoop is written in Java and is part of
the Apache Software Foundation.  It is a fairly mature piece of software
and is used in production at numerous companies. \cite{hadoop}

MapReduce works by splitting the input, which commonly is in the size
range of many gigabytes, into n parts and lets each node of the computing
cluster work on one or more parts. The mapper on each node emits key-value
pairs which are fed into the reduce phase. The final output is collated,
usually on a distributed file system.

Analysing the massive amounts of log data raises high demand for computing
power. The MapReduce \cite{mapreduce} pattern suits the analysing issue
well. The great benefit of MapReduce is that it runs on a computer cluster
of commodity hardware where computation takes place independently in
parallel. All results computed by each node are then reduced into one
single result.

Basically MapReduce is a two phase computation. First the input data is
partitioned and each of these partitions is sent to a \textit{mapper} which
performs computations on each row of the data. The mapper then outputs each
resulting computation on the format $\langle$\textit{key,value}$\rangle$. These
key-value pairs are received by a \textit{reducer} which performs the second
stage of the computation.
All Zohmg import programs are executed on Hadoop.

\subsection{Hadoop Distributed File System}

The goal of HDFS is to store large data sets while being able to handle
hardware failure and being deployed in heterogenous hardware and software
eco systems. Additionally the Hadoop and HDFS interaction is designed with
the notion that it is cheaper to move the computation, MapReduce programs,
to the data than the other way around.

\subsection{Hadoop streaming}

\subsection{Dumbo}

Dumbo is a Python framework for writing Hadoop programs in languages other than
Java, in a so-called Streaming Mode. It is used extensively at Last.fm for
writing short prototypes. \cite{dumbo}

All Zohmg programs written in Python use Dumbo to execute on Hadoop.

HBase is a free software implementation of Google's BigTable data store
described in \cite{bigtable}. It is a sub-project of Hadoop, written in
Java and a part of the Apache Software Foundation. HBase is still in its
infancy and few organizations use it for mission-critical applications.
\cite{hbase}

BigTable is a data store designed for traversal of very large data sets.
Time-stamped cells inside column-familes in rows. The row has a key and
any number of cells attached to it. The rows in a table are sorted by the
row key.  The typical use-case of BigTable is to scan a range of rows. It
is therefor important to set up one's keys so that related data is close
to each other. For example, if it makes sense to traverse the data in
chronological order, the key might contain a representation of the data's
timestamp. The classic example is to have the reversed domain name (i.e.
com.google.www) as the row key, which means that all sub-domains of any
domain are next to each other.

All Zohmg data is stored in HBase.

The data model is a simple one: Rows of data are stored in named tables.
Each row consists of a sortable row key and an arbitrary number of columns.
The columns are of the form "column-family:qualifier", where the column
family is one of a number of fixed column families defined by the table,
and the qualifier is an arbitrary string specified by the application. In
effect, column families are pre-announced while qualifiers are not. The
contents of each column family are stored together, so the user will want
to store items that have similar characteristics in the same column family.
Each data point is called a cell. A cell is an uninterpreted array of
bytes---there are no data types, and it is up to the application to
interpret the data correctly. The table can store any number of timestamped
versions of each cell, allowing versioning of data.

The model above can indeed be thought of as a sparse distributed
multidimensional sorted map. In concept, this multidimensional map is
identical to a nested hash map from your favourite programming language. The
dimensions of this multidimensional map is mapped out on the row key, the
column family and the column qualifier, and possibly also of the version
timestamp.

@@

@@

Sorted distributed multidimensional sparse hashtable.

\subsubsection*{Rows}

Each row in a BigTable is associated with a unique key. All the data is
stored sorted by rowkey. Because of this it is important to have a thorough
analysis of how the data is going to be accessed in order to store it
efficiently.

\subsubsection*{Regions}

BigTable use the concept of tablets, HBase in turn uses regions. In order
to be consistent with the rest of this paper we henceforth use the term
region.

A region is a partition of the data stored on a node in the computer
cluster. A region server (referred to as tablet server in \cite{bigtable})
holds information on which regions are located on which nodes. More
precisely the region server knows what range of rowkeys are in each region
on each node. This behaviour makes it fast to access data while keeping
reliability and scalability features.

% TODO> why is this Tool & BigTable-subsection called Dimensions?
% Disambiguation needed. -fredrik
\subsection*{Dimensions}

The degrees of freedom in BigTable.

\subsubsection*{Column-families}

Each column-family is stored as a single map-file in HBase.

\subsubsection*{Column-qualifiers}

Each column-family is further divided into one or more column-qualifiers.

@@

Depending on how the rowkeys are formatted different hierarchies are
created. The order in which the data is stored is important, it defines how
effective scanning data will be. In the case of having thousand rows and
requesting ten out of these, then only one por cent of the data is
interesting. If a scan would be required to visit all the thousand rows a
lot of rows are skipped. Skipping rows is expensive in that sense that they
are but the result is thrown away. The goal is to push this cost to a bare
minimum, using as many of the rows as possible.

The minimum cost of scanning data is achieved by ordering the data so that
a scan uses every visited row. In the case with thousand rows of which ten
are wanted, would mean that ten rows scan are scanned of which all are
used.
\subsection{Thrift}

A remote procedure call framework for building scalable cross-language
services. Thrift combines a software stack with a code generation engine to
build services that work seamlessly between a wide range of programming and
scripting languages. \cite{thrift}

Thrift was originally developed at Facebook and in April 2007 it was
released as free software. It entered the Apache Incubator in May 2008.

@@

\subsection{JSON}

JSON\footnote{JavaScript Object Notation} is a lightweight text-based
human-readable data interchange format. It is used for representing simple
data structures and associative arrays (called objects). JSON is often used
for serializing structured data and transmitting it over network
connection.

\subsection{YAML}

YAML\footnote{YAML Ain't a Markup Language} is a human-readable data
serialization format available for a wide range of programming and
scripting languages.

The outline and markup of YAML documents makes it well suited for data
which humans are likely to view or edit, such as configuration files.

\section{Java}

Both Hadoop and HBase are written in Java---a high-level object-oriented
programming language.

Most Hadoop users write their MapReduce programs in Java, since this is the
native language of Hadoop. \cite{java}

\subsection{PyYAML}

PyYAML is a complete YAML 1.1 compliant parser and emitter for Python.
\cite{pyyaml}

\subsection{Setuptools}

Python Setuptools is a framework---built on top of the Python module
distutils---for basic package management of Python software. Amongst other
things it creates and installs Python Eggs\footnote{A Zip file which
bundles Python software files.}. \cite{setuptools}

\subsection{simplejson}

A simple, fast, complete and extensible JSON encoder and decoder for
Python. The Python module simplejson is implemented entirely in Python and
has no external dependencies. However a C extension which provides a
serious speed boost is included in simplejson. \cite{simplejson}

Data sets and environment.

@@

Zohmg is written mostly in Python. Zohmg uses the following Python modules
internally; it depends on them to be available on the system.

@@

TODO: More general about middleware, move Paste mentioning to Implementation
chapter.

@@

% COMMENT: do we really need to repeat this?
As stated in chapter 3, Zohmg is built upon existing tools and software. It
uses Dumbo to run jobs on Apache Hadoop, and uses Apache HBase to persist the
data.

@@

\begin{itemize}
\item \texttt{Serializing}
Serializing data is the task of converting an object into a sequence of
bits so that it can be stored on a storage medium or transmitted over a
network link. Rereading the sequence of bits should restore the original
state of the object it was in before the serialization.
\end{itemize}

>>>>

% chapterquote
\newcommand{\chapterquote}[2] {
\begin{quote}
\textit{"{#1}"}

@@

At Last.fm there is a central repository where metrics and statistics for
various services are stored. There is a web site for internal use that
displays graphs of the services and their statistics. Examples of statistics
are web site usage broken down by country, scrobbling broken down by client
and daily subscription counts.

@@

% TODO: Why do we do these avgränsningar.

@@

OLAP is something. It's drawbacks are unknown.

@@

Mention data sets here.

@@

@@

The slice operation fixes the values of one or several dimensions in the cube.
The data omitted from this operation would be any data associated with the
values of the dimension that were not fixed. The dice operation is a slice on
more than two dimensions.

@@

TODO: Time series is one of the most important measures in log analyzing, it
is also very natural when the data is time-stamped. \cite{discoveringweb}

@@

\section{Techniques and tools}

TODO: Why do we chose HBase in favour of other techniques (i.e. SQL)?

@@

\section{Procedure}

TODO: Explain our working procedure regarding
\begin{itemize}
\item[-] Exploration
\item[-] Prototyping
\item[-] Real-world data
\item[-] Litterature studies
\item[-] Design phase
\end{itemize}

Hadoop is a free software implementation of Google's MapReduce-framework; it is
written in Java and is part of the Apache Software Foundation. Hadoop is a
fairly mature piece of software and is used in production at numerous companies,
among them Last.fm.
\cite{hadoop}

MapReduce works by splitting the input, which commonly is in the size
range of many gigabytes, into \textit{n} parts and lets each node of the
computing cluster work on one or more parts. The mapper on each node emits
key-value pairs which are fed into the reduce phase. The final output is
collated, usually on a distributed file system.

\subsection*{Hadoop Distributed File System}

The goal of HDFS is to store large data sets while being able to handle
hardware failure and being deployed in heterogenous hardware and software
eco systems. Additionally, the Hadoop and HDFS interaction is designed with
the notion that it is cheaper to move the computation, MapReduce programs,
to the data than the other way around.

\subsection*{Hadoop streaming}

\subsubsection*{Dumbo}

Dumbo is a Python framework for writing Hadoop programs in Python; it exploits
the Hadoop Streaming Mode to do so. It is used extensively at Last.fm for
writing short prototypes. \cite{dumbo}

HBase is a free software implementation of Google's BigTable. \cite{bigtable} It
is a sub-project of Hadoop, written in Java and a part of the Apache Software
Foundation. HBase is still in its infancy and few organizations use it for
mission-critical applications. \cite{hbase}

BigTable is a data store designed for traversal of very large data sets.
Time-stamped cells inside column-families in rows. The row has a key and any
number of cells attached to it. The rows in a table are sorted by the key.
The typical use-case of BigTable is to scan a range of rows. It is therefor
important to set up the row keys so that related data is close to each other.
For example, if it makes sense to traverse the data in chronological order, the
key might contain a representation of the data's timestamp. The classic example
is to have the reversed domain name (i.e. com.google.www) as the row key, which
means that all sub-domains of any domain are next to each other.

@@

The data model is a simple one: Rows of data are stored in labeled tables. Each
row is indexed with a row key, a column key, and a timestamp.  The columns are
of the form "column-family:qualifier", where the column-family is one of a
number of fixed column-families defined by the table, and the qualifier is an
arbitrary string specified by the application. In effect, column families are
pre-announced when creating the table while qualifiers are not. The contents of
each column family are stored together, so the user will want to store items
that have similar characteristics in the same column family.  Each data point is
called a cell. A cell is an uninterpreted array of bytes, there are no data
types, and it is up to the application to interpret the data correctly. The
table can store any number of timestamped versions of each cell, allowing
versioning of data.

\begin{center}
    \begin{table}[h]
    \begin{tabular}{|l|l|l|l|l|}
    \hline
    \small \bf Row Key & \small \bf Time & \small \bf Column &
    \multicolumn{2}{|c|}{\small \bf Column} \\
     & \small \bf Stamp & \small "contents:" &
    \multicolumn{2}{|c|}{\small "anchor:"} \\
    \hline
     & t9 & & \small "anchor:cnnsi.com" & "CNN" \\
     & t8 & & \small "anchor:my.lok.ca" & "CNN.com" \\
    "com.cnn.www" & t6 & \small "{\textless}html\textgreater..." & & \\
     & t5 & \small "{\textless}html\textgreater..." & & \\
     & t3 & \small "{\textless}html\textgreater..." & & \\
    \hline
    \end{tabular}
    \caption{Example HBase table.}
    \end{table}
\end{center}

\vspace{-24pt}

\noindent The model above can indeed be thought of as a sparse distributed
multidimensional sorted map. In concept, this multidimensional map is identical
to a nested hash map. The dimensions of this multidimensional map is mapped out
on the row key, the column family and the column qualifier, and possibly also of
the version timestamp.

% TODO> why is this Tool & BigTable-subsection called Dimensions?
% Disambiguation needed. -fredrik
\subsection*{Physical storage}

Each column-family is stored as a single file in HBase; it is then subdivided
into column-qualifiers within the same file. This is important to consider when
designing application and schemas, since reading from several column-families
will open several files.

\subsection*{Sparsity}

When many cells are empty, the table is said to be sparse. This can potentially
be a problem, because even sparse cells may cost storage space.  BigTable and
HBase, however, don't store empty cells; as consequence sparsity is not
associated with a great storage cost.
\cite{olap_data_scalability,olap_solutions}

@@

Depending on how the row keys are formatted different, hierarchies are created.
The order in which the data is stored is important, it defines how effective
scanning data will be. Sparsity also comes into play concerning effective
scanning. In the case of having thousand rows and requesting ten out of these,
then only one por cent of the data is interesting. If a scan would be required
to visit all the thousand rows a lot of rows are skipped. Skipping rows is
expensive in that sense that they are but the result is thrown away. The goal is
to push this cost to a bare minimum, using as many of the rows as possible.

The minimum cost of scanning data is achieved by ordering the data so that
a scan uses every visited row. In the case with thousand rows of which ten
are wanted, would mean that ten rows scan are scanned of which all are
used.

\subsection*{Thrift}

@@

Thrift is a remote procedure call framework for building scalable cross-language
services. Thrift combines a software stack with a code generation engine to
build services that work seamlessly between a wide range of programming and
scripting languages.

Thrift was originally developed at Facebook and in April 2007 it was released as
free software. It entered the Apache Incubator in May 2008. \cite{thrift}

\section{Dependability}

Regarding the infrastructure, both Hadoop and HBase have high availability,
reliability, and scalability.

Both Hadoop and HBase are distributed systems; making them highly available out
of the box. Although the Job Tracker\footnote{All client applications are
submitted to Hadoop's Job Tracker.} is a single point of failure for Hadoop.
The HDFS NameNode\footnote{A unique server which keeps the directory tree of all
files in HDFS.} is also a single point of failure. However, a secondary HDFS
NameNode, which regularly takes a snapshot of the primary NameNode's state, is
spawned. There exist no automatic fail-over though, if the main the primary
NameNode crashes manual intervention is needed.

Both systems have high reliability. Data on HDFS are replicated to---typically
three---other nodes in the cluster. Jobs submitted to Hadoop also have a high
reliability: If a node does not succeed with a job it is rescheduled to another
free node which executes it.

Regarding scalability Hadoop and HBase scales close to linearly when nodes are
added to the cluster.

While relational databases can be made highly available and reliable with load
balancing and so called \textit{Sharding}, the setup is non-trivial. Regarding
scalability, relational databases are highly dependent on having data in working
memory; thus they scale poorly.
% TODO: [citation needed]

@@

Because there existed a Python module for writing MapReduce programs (Dumbo),
Python was a natural choice during the design and exploration phases. The main
reason was the rapid development of Python programs, several times shorter than
their Java counter parts made them faster to develop.

The Zohmg needs configuration of the user's data sets and the environment for
Hadoop and Dumbo.

The data sets are configured using YAML\footnote{YAML Ain't a Markup Language},
a human-readable data serializaton format. Any human-readable configure file
method could have been chosen, e.g. INI files\footnote{A simple configuration
mechanims where one would have lines of \texttt{variable = value}.}. YAML was
choosen because it is is very simple for humans to read and edit, and likewise
for machines.

The environment is a Python script which defines run-time variables, required by
Hadoop and Dumbo. Using a Python scripts for run-time variables leaves the
problem of validation to Python.

@@

In order to join together all the different abilities which were wanted from the
data server, it was designed as a middleware application. Since the main goal
was not to build a middleware framework from scratch, an existing such was used.

The data export server is built with a WSGI\footnote{Web Server Gateway
Interface} framework, it includes both middleware and application parts. The
middleware applications are responsible for dispatching requests to the
corresponding application or file.

When a user queries the data server, JSON\footnote{JavaScript Object Notation}
is returned. JSON was chosen because it is i simple text serialization format
which is widely used in web development.

@@

@@

\begin{tabular}[t]{lp{100mm}}
\textbf{JSON} &
JavaScript Object Notation, a lightweight text-based human-readable data
interchange format. It is used for representing simple data structures and
associative arrays (called objects). JSON is often used for serializing
structured data and transmitting it over network connection.

 \\

\textbf{Serializing} &
Serializing data is the task of converting an object into a sequence of bits,
making it possible to store on a storage medium or transmitted over a network
link. Rereading the sequence of bits should restore the original state of the
object it was in before the serialization. \\
\end{tabular}

\chapter{Text with no home and no current purpose}
